ASTR480 Final Report:\\ \textit{Asteroids in TESS}

### 2024-10-20 FinalReport/FinalReportDraft.tex
Final Report Done

\usepackage{chngcntr}

@@

% %*Background
The Transiting Exoplanet Survey Satellite (TESS) is a wealth of high-cadence photometric data. 
Asteroids move across the TESS field, and their intrinsic rotation varies their brightness periodically. 
Here, I present an analysis of the asteroids in TESS sector 29, as an addition to the \texttt{TESSELLATE} pipeline's analysis.

@@

\subsection{Asteroids}\label{SubSec:Asteroid}

@@

% The semi-major axis, $a$, is the largest distance from the sun that the asteroid achieves on its elliptical orbit.
% Main belt asteroids have a typical $a$ of \qtyrange{2}{3.3}{\au} \citep{DeMeo2015}.
% The bounds between the inner, middle and outer main belt at \qty{2.5}{\au} and \qty{2.82}{\au}, these separations are due to resonances with Jupiter (the 3:1 and 5:2 mean motion resonances mark the edges of middle main belt) and Saturn (the $\nu_6$ secular resonance cuts off the lowest $a$ edge of the inner belt) carving swathes where asteroids cannot be in stable orbits.

% The eccentricity, $e$, of this orbit is the next proper element. $e=0$ implies a perfectly circular orbit, $0<e<1$ are ellipses, $e=1$ give a parabolic orbit, and $e>1$ are unbound, hyperbolic orbits.
% Main belt asteroids have small $e$, in the range of  \citep{DeMeo2015}
% Comets have larger $e$, with some long period comets having $e\approx 1$, but the typical range is \qtyrange{0.2}{0.7}{} \citep{Lewis2012}.
% Interstellar objects (ISOs) have $e>1$ as they are on unbound hyperbolic trajectories.

% The inclination, $i$, of asteroids is also a key orbital element, it is the angle relative to the ecliptic plane.
% In principle asteroids can span the whole range of inclinations, though they rarely do.
% Main belt asteroids have typical inclinations of \qtyrange{0}{30}{\degree} \citep{DeMeo2015}, so the TESS cameras closer to the ecliptic plane are expected to have the most asteroids in their fields.

@@

%TODO Rotation of asteroids
As an asteroid rotates, the amount of sunlight reflected changes with time.
This produces a lightcurve that is often sinusoidal in appearance.
These lightcurves are normally ``double-peaked'', with two maximums for one full rotation, due to symmetries in the object's shape. 
The shape of the asteroid has a large effect on the scale of variation in the lightcurve \citep{Durech2015}, the more elongated the object is, the larger the variation.
The rate of rotation is limited by the size of the asteroid \citep{Pravec2000}, as the body's internal strength is placed under too much strain if large
asteroids rotate with short periods.

@@

\subsection{TESS}\label{SubSec:TESS}

%Why TESS
TESS is a large area, high-cadence imaging, space telescope  \citep{Ricker2014}.
TESS is tasked with observing one piece of sky for \qty{27}{\day} at a time, which defines a sector.
The telescope delivers \qtyproduct{96 x 24}{\degree} full frame images (FFIs).
These FFIs are built from stacked \qty{2}{\second} exposures, leveraging the short readout times of the 16 charged-coupled devices (CCDs) spread over four cameras.
With the initial cadence for these full frame images set to \qty{30}{\minute}, the time resolution of TESS is unparalleled.
As the mission was extended after TESS had already mapped the entire sky, the length of the FFIs decreased to \qty{10}{\minute} and then later to \qty{200}{\second}.
This high time resolution and observation area does come at the cost of spatial resolution, as the pixels are each \qty{21}{\arcsec} square.

@@

\subsection{Asteroid Properties}\label{SubSec:AstPropRes} %TODO fix name

@@

\subsection{Rotation Period Analysis Results}\label{SubSec:PerRes}

@@

The distribution of the amplitude of the lightcurves is given in \autoref{Fig:ampDistr}.
The peak of the distribution at $\frac13$ \unit{\mag} and is right skewed.
The high $\Delta m$ values are from lightcurves of dubious quality.
Manually checking the analysis, it is clear that these asteroids above $\Delta m =\qty{1.5}{\mag}$ have all been affected contamination in their lightcurves, and have untrustworth photometry.
Some examples are given in the appendix, Figures \ref{ApFig:Takashimizuno} \ref{ApFig:Wuyeesun} and \ref{ApFig:Locarno}

@@

Lomb-Scargle periodograms can deal with large gaps in the time data, but would require the flux values to have the same normalisation.
\texttt{TESSELLATE} is self-consistent in the data reduction for a sector, but comparing fluxes between sectors may have some difficulties.
By assuming the mean flux of the asteroid should be physically constant, subtracting this off the lightcurves before the periodogram is run, as is already done, will start to mitigate this difference. 
There are also phase and viewing angle effects on the lightcurve that are important to correct for with large time gaps. %????
These were not considered for the single sector analysis because the change in these values is not significant over a month.

@@

A Lomb-Scargle periodogram was used to detect the periods of each object. 
From the model that \texttt{Astropy} determined was the best fit to the data, a $\Delta m$ was calculated to quantify the amplitude of the lightcurve. 
The distributions of these broadly agree with the literature.

Of the 5664 objects with $V<\qty{20}{\mag}$ in found in sector 29, the rotation properties of 374 asteroids are calculated. 
These are the objects that have met all the requirements placed on their lightcurve and periodogram, and the checks were found to be \qty{70}{\percent} accurate.
59 of these asteroids have rotation periods documented in the LCDB already \citep{Warner2009}.
315 new asteroid rotation periods have been determined, as well as the amplitude of the variation in the lightcurve, \qty{80}{\percent} of the total asteroids that pass the checks.

Some bodies appear to have more extreme variation of their lightcurves that the unusual \omuamuans.
These were checked and have contamination in the lightcurves. 
This contamination has managed to pass undetected through sigma-clipping and other quality checks on the periodogram. 
There is evidently a slight issue with my checks

I have shown that \texttt{TESSELLATE} detects asteroids far less often than they appear in the FFIs.
Most objects are matches $<100$ times, while the forced photometry lightcurves average more than 2000 observations. 
Those that are matched to TESSELLATE detections can be filtered out of further transient searches. 
The forced photometry detects more than \qty{50}{\percent} of asteroids brighter than

@@

Here I present some more periodogram examples, Figures \ref{ApFig:Ceres} \ref{ApFig:Takashimizuno}, \ref{ApFig:Wuyeesun} and \ref{ApFig:Locarno}.
This provides context to what I claim in text especially about the contamination of the lightcurves.
As they are all so similar to \autoref{Fig:PeriodEx} and quite large, so I do not place them where they are referred to.

@@

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{Figures/Periodogram1999 VN44Resid.pdf}
  \caption[(36240) 1999 VN44 Periodogram]{Same as \autoref{Fig:PeriodEx}, but for (36240) 1999 VN44.
  This object was in the data of both \citet{Pal2020} and \citet{McNeill2023}, but the calculated period here was not the same.}
  \label{ApFig:1999VN44}
\end{figure}

>>>>

@@

% %*Background
The Transiting Exoplanet Survey Satellite (TESS) is a wealth of high-cadence photometric data. 
Asteroids move across the TESS field, and their intrinsic rotation varies their brightness periodically. 
Here, an analysis of the asteroids in TESS sector 29 is presented, as an addition to the \texttt{TESSELLATE} pipeline's analysis.

@@

\subsection{Asteroids and their Rotation Properties}\label{SubSec:Asteroid}

@@

@@

As an asteroid rotates, the amount of sunlight reflected changes with time.
This produces a lightcurve that is often sinusoidal in appearance.
These lightcurves are normally ``double-peaked'', with two maximums for one full rotation, due to symmetries in the object's shape. 
The shape of the asteroid has a large effect on the scale of variation in the lightcurve \citep{Durech2015}, the more elongated the object is, the larger the variation.
The rate of rotation is limited by the size of the asteroid \citep{Pravec2000}, as the body's internal strength is placed under too much strain if large
asteroids rotate with short periods.

@@

\subsection{TESS and its History of Asteroid Analysis}\label{SubSec:TESS}

TESS is a large area, high-cadence imaging, space telescope  \citep{Ricker2014}.
TESS is tasked with observing one piece of sky for \qty{27}{\day} at a time, which defines a sector.
The telescope delivers \qtyproduct{96 x 24}{\degree} full frame images (FFIs).
These FFIs are built from stacked \qty{2}{\second} exposures, leveraging the short readout times of the 16 charged-coupled devices (CCDs) spread over four cameras.
With the initial cadence for these full frame images set to \qty{30}{\minute}, the time resolution of TESS is unparalleled.
As the mission was extended after TESS had already mapped the entire sky, the length of the FFIs decreased to \qty{10}{\minute} and then later to \qty{200}{\second}.
This high time resolution and observation area does come at the cost of spatial resolution, as the pixels are each \qty{21}{\arcsec} square.

@@

\subsection{The Distribution of Asteroid Properties}\label{SubSec:AstPropRes}

@@

\subsection{Rotation Period and Amplitude Variation Results}\label{SubSec:PerRes}

@@

The distribution of the amplitude of the lightcurves is given in \autoref{Fig:ampDistr}.
The peak of the distribution at $\frac13$ \unit{\mag} and is right skewed.
The high $\Delta m$ values are from lightcurves of dubious quality.
Manually checking the analysis, it is clear that these asteroids above $\Delta m =\qty{1.5}{\mag}$ have all been affected contamination in their lightcurves, and have untrustworthy photometry.
Some examples are given in the appendix, Figures \ref{ApFig:Takashimizuno} \ref{ApFig:Wuyeesun} and \ref{ApFig:Locarno}

@@

Lomb-Scargle periodograms can deal with large gaps in the time data, but would require the flux values to have the same normalisation.
\texttt{TESSELLATE} is self-consistent in the data reduction for a sector, but comparing fluxes between sectors may have some difficulties.
There are also phase and viewing angle effects on the lightcurve that are important to correct for with large time gaps. %????
These were not considered for the single sector analysis because the change in these values is not significant over a month.

@@

A Lomb-Scargle periodogram was used to detect the periods of each object. 
From the model that \texttt{Astropy} determined was the best fit to the data, a $\Delta m$ was calculated to quantify the amplitude of the lightcurve. 
The distributions of these properties broadly agree with the literature.

Of the 5664 objects with $V<\qty{20}{\mag}$ in found in sector 29, the rotation properties of 374 asteroids are calculated. 
These are the objects that have met all the requirements placed on their lightcurve and periodogram, and the checks were found to be \qty{70}{\percent} reliable.
59 of these asteroids have rotation periods documented in the LCDB already \citep{Warner2009}.
So 315 new asteroid rotation periods have been determined, as well as the amplitude of the variation in the lightcurve, \qty{80}{\percent} of the total asteroids that pass the checks.

Some bodies appear to have more extreme variation of their lightcurves than the unusual \omuamuans.
These were checked and have contamination in the lightcurves. 
This contamination has managed to pass undetected through sigma-clipping and other quality checks on the periodogram. 
There is evidently a issue with my checks, and more data will help to refine them.

I have shown that \texttt{TESSELLATE} detects asteroids far less often than they appear in the FFIs.
Most objects are matches $<100$ times, while the forced photometry lightcurves average more than 2000 observations. 
Those that are matched to TESSELLATE detections can be filtered out of further transient searches. 
The forced photometry detects more than \qty{50}{\percent} of asteroids brighter than $V = \qty{16.36(0.73)}{\mag}$ from analysis of the recovery fraction.
This agrees with other work on the detection efficiency of \texttt{TESSELLATE}

@@

Here I present some more periodogram examples, Figures \ref{ApFig:Ceres} \ref{ApFig:Takashimizuno}, \ref{ApFig:Wuyeesun} \ref{ApFig:Locarno}, and \ref{ApFig:1999VN44}.
This provides context to what I claim in the text, especially about the contamination of the lightcurves.
As they are all so similar to \autoref{Fig:PeriodEx} and quite large, so I do not place them where they are referred to.

@@

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{Figures/Periodogram1999 VN44Resid.pdf}
  \caption[(36240) 1999 VN44 Periodogram]{Same as \autoref{Fig:PeriodEx}, but for (36240) 1999 VN44.
  This object was in the data of both \citet{Pal2020} and \citet{McNeill2023}, but period the calculated here was not the same, due to the low mean flux.}
  \label{ApFig:1999VN44}
\end{figure}
